---
title: "p8105_hw2_at3583"
author: "Alexandria D.Taylor,"
date: "September 28, 2025"
output: github_document
---

library(tidyverse)
library(janitor)
library(readr)

# Problem 1 

```{r}
library(readr)
# Import .csv files for FiveThirtyEight data
pols_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/pols-month.csv")
unemp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/unemployment.csv")
snp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/snp.csv")
```

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
# clean pols_df using separate
pols_df <- pols_df %>%
separate(mon, into = c("year", "month", "day"), sep ="-") %>%
mutate(across(c(year, month, day), as.integer)) %>%
mutate(
month = month.name[month],
president = ifelse(prez_gop == 1, "gop", "dem")
) %>%
select(-prez_dem, -prez_gop, -day)

# clean snp_df using separate
snp_df <- snp_df %>%
separate(date, into = c("month", "day", "year"), sep = "/") %>%
mutate(across(c(year, month, day), as.integer)) %>%
mutate(month = month.name[month]) %>%
select(year, month, everything(), -day) %>%
arrange(year, month)

#clean unemp_df from wide to longer format using tidy
unemp_df <- unemp_df %>%
pivot_longer(
Jan:Dec,
names_to = "month",
values_to = "unemployment"
) %>%
rename (year = Year) %>%
mutate(
  month = str_to_title(month),
  year = as.integer(year),
  month = month.name[match(str_to_title(month), month.abb)]
)
  
# Merge pols, snp and unemp dataframes
merged_df <- pols_df %>%
  left_join(snp_df, by = c("year", "month")) %>%
  left_join(unemp_df, by = c("year", "month"))
```

## The **pols_df** dataset contains information on national politicians by party, including `r ncol(pols_df) - 3` variables on party composition and affiliation of the sitting president, while the **snp_df** dataset contains monthly closing values of the S&P stock index, with `r nrow(snp_df)` observations. The **unemployment_df** dataset reports U.S. unemployment rates by year and month, tidied to long format with `r ncol(unemp_df)` fields. After cleaning and merging of all three data frames,the resulting datasets contains **`r nrow(merged_df)` observations** and **`r ncol(merged_df)` variables**, spanning the years **`r min(merged_df$year)` to `r max(merged_df$year)`**. Highlighted variables include `year`, `month`, the `president` indicator (with values `r paste(unique(merged_df$president), collapse = ", ")`), party composition counts such as `gov_gop`, `sen_gop`, `rep_gop`, the S&P closing index (`close`), and unemployment rates (`unemployment`).  

# Problem 2


