---
title: "p8105_hw2_at3583"
author: "Alexandria D.Taylor,"
date: "September 28, 2025"
output: github_document
---


# Problem 1 

library(tidyverse)
library(janitor)
library(readr)

# Import .csv files for FiveThirtyEight data
pols_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/pols-month.csv")
unemp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/unemployment.csv")
snp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/snp.csv")


```{r}
library(dplyr)
library(tidyr)
pols_df <- pols_df %>%
#Clean pols dataframe and others using use separate()
pols_df = pols_df %>%
  clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep ="-") %>%
  mutate(across(c(year, month, day), as.integer)) %>%
  mutate(
    month = month.name[month],
    president = ifelse(prez_gop == 1, "gop", "dem")
    ) %>%
  select(-c(prez_dem, prez_gop, day))
  select(-prez_dem, -prez_gop, -day)

#Clean snp dataframe
snp_df <- snp_df %>%
  clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(across(c(year, month, day), as.integer)) %>%
  mutate(month = month.name[month]) %>%
  select(year, month, everything(), -day) %>%
  arrange(year, month)

#Clean unemp_df
unemp_df <- unemp_df %>%
  clean_names() %>%
  pivot_longer(
   jan:dec,
    names_to = "month",
    values_to = "unemployment"
  ) %>%
  mutate(
    month = str_to_title(month),
    year = as.integer(Year)
  ) %>%
  select(year, month, unemployment)
```

```{r}
# Merge pols, snp and unemp dataframes
merged_df <- pols_df %>%
  left_join(snp_df, by = c("year", "month")) %>%
  left_join(unemp_df, by = c("year", "month"))
```

Summary: 

# Problem 2

