---
title: "p8105_hw2_at3583"
author: "Alexandria D.Taylor,"
date: "September 28, 2025"
output: github_document
---

# Problem 1 - FiveThirtyEight

```{r}
library(readr)
library(tidyverse)

# Import .csv files for FiveThirtyEight datasets
pols_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/pols-month.csv")
unemp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/unemployment.csv")
snp_df <- read_csv("/Users/goddid333/Desktop/GIT Items/fivethirtyeight_datasets/snp.csv")
```

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)

#clean pols_df using separate
pols_df <- pols_df %>%
separate(mon, into = c("year", "month", "day"), sep ="-") %>%
mutate(across(c(year, month, day), as.integer)) %>%
mutate(
month = month.name[month],
president = ifelse(prez_gop == 1, "gop", "dem")
) %>%
select(-prez_dem, -prez_gop, -day)

#clean snp_df using separate
snp_df <- snp_df %>%
separate(date, into = c("month", "day", "year"), sep = "/") %>%
mutate(across(c(year, month, day), as.integer)) %>%
mutate(month = month.name[month]) %>%
select(year, month, everything(), -day) %>%
arrange(year, month)

#clean unemp_df from wide to longer format using tidy
unemp_df <- unemp_df %>%
pivot_longer(
Jan:Dec,
names_to = "month",
values_to = "unemployment"
) %>%
rename (year = Year) %>%
mutate(
  month = str_to_title(month),
  year = as.integer(year),
  month = month.name[match(str_to_title(month), month.abb)]
)
  
# merge pols, snp and unemp dataframes
merged_df <- pols_df %>%
  left_join(snp_df, by = c("year", "month")) %>%
  left_join(unemp_df, by = c("year", "month"))
```

The **pols_df** dataset contains information on national politicians by party, including `r ncol(pols_df) - 3` variables on party composition and affiliation of the sitting president, while the **snp_df** dataset contains monthly closing values of the S&P stock index, with `r nrow(snp_df)` observations. The **unemployment_df** dataset reports U.S. unemployment rates by year and month, tidied to long format with `r ncol(unemp_df)` fields. After cleaning and merging of all three data frames,the resulting datasets contains **`r nrow(merged_df)` observations** and **`r ncol(merged_df)` variables**, spanning the years **`r min(merged_df$year)` to `r max(merged_df$year)`**. Highlighted variables include `year`, `month`, the `president` indicator (with values `r paste(unique(merged_df$president), collapse = ", ")`), party composition counts such as `gov_gop`, `sen_gop`, `rep_gop`, the S&P closing index (`close`), and unemployment rates (`unemployment`).  

# Problem 2 - Trash Wheel

```{r}
library(readxl)
library(janitor)
library(dplyr)
library(readr)

# read and clean Mr. Trash Wheel sheet
mr_trash <- read_excel("~/Desktop/GIT Items/202509 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    year = as.integer(readr::parse_number(year)),
    sports_balls = as.integer(round(sports_balls, 0)),
    wheel = "Mr. Trash Wheel"
  )

# read and clean Professor Trash Wheel sheet
prof_trash <- read_excel("~/Desktop/GIT Items/202509 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    year = as.integer(readr::parse_number(as.character(year))),
    wheel = "Professor Trash Wheel")

# read and clean Gwynns Falls Trash Wheel sheet
gwynnda <- read_excel("~/Desktop/GIT Items/202509 Trash Wheel Collection Data.xlsx", sheet = "Gwynns Falls Trash Wheel", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    year = as.integer(readr::parse_number(as.character(year))),
    wheel = "Gwynnda")
```

```{r check structure,include=FALSE}
# check data structure before binding
str(mr_trash$year)
str(prof_trash$year)
str(gwynnda$year)
```

```{r}
# combine dataframes into one tidy dataset
trash_df <- bind_rows(mr_trash, prof_trash, gwynnda)
```

 The combined Trash Wheel dataset contains `r nrow(trash_df)` observations and `r ncol(trash_df)` variables, collected across `r min(trash_df$year, na.rm = TRUE)` to `r max(trash_df$year, na.rm = TRUE)`. Variables include collection date (`year`, `month`), volume and weight of trash collected (`volume_cubic_yards`), (`weight_tons`), and the counts of items such as `plastic_bottles`, `glass_bottles`, and `cigarette_butts`. Overall, Professor Trash Wheel has collected a total of `r sum(trash_df$weight_tons[trash_df$wheel == "Professor Trash Wheel"], na.rm = TRUE)` tons of trash,while Gwynnda collected `r sum(trash_df$cigarette_butts[trash_df$wheel == "Gwynnda" & trash_df$month == "June" & trash_df$year == 2022], na.rm = TRUE)` cigarette butts in June 2022.
